---
title: "Refactoring a multi-dataset ggplot2 chain"
subtitle: "The same plot, refactored twice, changing only Decision C"
format:
  gfm:
    toc: true
    fig-width: 7
    fig-height: 4.5
execute:
  warning: false
  message: false
---

A worked example for the `refactor-ggplot-helper` skill, covering the case the
ggtilecal example does not: a plot whose layers inherit nothing. See
[README.md](README.md) for the findings; this document renders the plots.

Reproduce with `install.packages(c("xmap", "ggforce", "ggrepel"))`.

```{r}
#| label: setup
library(xmap)
library(dplyr)
library(ggplot2)

xm <- as_xmap_tbl(demo$simple_links, xcode, alphacode, weight)
xm
```

## Before

The original one-off chain. Note the **empty** `ggplot()`, the four separate
datasets, and the four `aes()` blocks each carrying its own `data =`. Nothing
inherits.

```{r}
#| label: fig-before
#| fig-cap: "Before: a one-off multi-dataset ggplot2 chain"
#| file: before-bigraph.R
```

## After (a): layer-focused

Decision C → group list arguments by ggplot2 object type, the same shape as
`gg_facet_wrap_months()`.

```{r}
#| label: after-layer
#| file: after-layer-focused.R
#| fig-show: hide
```

```{r}
#| label: fig-after-layer
#| fig-cap: "After: layer-focused grouping"
gg_diagonal_bigraph(xm)
```

The signature:

```{r}
#| label: sig-layer
args(gg_diagonal_bigraph) |> str()
```

Still composable, and the prep is usable without plotting:

```{r}
#| label: fig-layer-compose
#| fig-cap: "Composition with `+` still works"
gg_diagonal_bigraph(xm) + theme(legend.position = "right")
```

```{r}
#| label: layout-standalone
str(calc_bigraph_layout(xm), max.level = 1)
```

### The removal-semantics footgun

Emptying `.scale_coord` builds without error, but silently de-synchronises the
colour and fill palettes so a link and its weight label stop sharing a colour —
and flips the diagram vertically:

```{r}
#| label: fig-removal
#| fig-cap: "`.scale_coord = list()` — builds fine, wrong plot"
gg_diagonal_bigraph(xm, .scale_coord = list())
```

## After (b): chart-focused

Decision C → group by what a reader of the chart sees. `.links` merges
`geom_diagonal()` with `scale_alpha_continuous()` across the geom/scale
boundary, because the alpha scale exists only to make edge opacity read as
weight.

```{r}
#| label: after-chart
#| file: after-chart-focused.R
#| fig-show: hide
```

```{r}
#| label: fig-after-chart
#| fig-cap: "After: chart-focused grouping"
gg_link_xmap_nodes(xm)
```

A scalar exposed off a fixed component — Decision B's third category, which
survives here but is voided under layer-focused grouping:

```{r}
#| label: fig-chart-scalar
#| fig-cap: "`node_fill` exposed as an ordinary scalar argument"
gg_link_xmap_nodes(xm, node_fill = "lightblue")
```

A user-supplied bare geom, acquiring its data and mapping via
`bind_layer_data()`:

```{r}
#| label: fig-chart-bare-geom
#| fig-cap: "A bare `geom_diagonal()` supplied by the user"
gg_link_xmap_nodes(xm, .links = list(ggforce::geom_diagonal(linewidth = 2)))
```

## Equivalence

Both refactors reproduce the original exactly:

```{r}
#| label: equivalence
b0 <- ggplot_build(plot_xmap_bigraph(xm))$data
b1 <- ggplot_build(gg_diagonal_bigraph(xm))$data
b2 <- ggplot_build(gg_link_xmap_nodes(xm))$data

c(layer_focused = isTRUE(all.equal(b0, b1)),
  chart_focused = isTRUE(all.equal(b0, b2)))
```
